---
title: Using survey weights and survey design variables
subtitle: R examples
institute: UK Data Service

date: "June 2026"
date-format: MMMM YYYY
format:
  html:
    css: ukds_long.css
    embed-resources: false
    toc: true
    toc-depth: 2
    number-sections: true

execute:
  echo: true
  warning: false
  message: false
---

# Introduction

Survey weights help make survey estimates more representative of the target population. Many surveys also use complex sample designs involving stratification and clustering. These design features should be taken into account when estimating standard errors and confidence intervals.

This resource shows how to apply survey weights and account for survey design in R through examples using data from the 2017 British Social Attitudes Survey.

The examples show how to:

- apply survey weights when calculating a mean.

- specify a survey design using the survey package.

- estimate means and proportions with confidence intervals.

- produce estimates for population subgroups.

For an introduction to the concepts covered here, including survey weights, stratification, clustering and design variables, see the companion resource *Survey Explainer: Survey weights and design variables*.

You can use this resource as a simple reference, or follow the examples using your own copy of the data. Information on the required software and data is provided below.

# Survey weighting and design in R

There are several ways to apply survey weights in R. A key distinction is between general weighting functions which apply weifhts without accounting for the survey design, and survey-specific approaches, which can account for survey design when estimating measures of precision such as confidence intervals.

## **General weighting functions**

Base R provides some functionality for weighted analysis, including functions such as `weighted.mean()` and regression models like `lm()` that allow the use of weights. Additional R packages offer convenient functions for calculating weighted estimates. For example, the `Hmisc` package includes functions such as `wtd.mean()`, `wtd.var(),` and `wtd.quantile()`, while the summarytools package supports weighted frequency tables and common descriptive statistics.

## **Survey analysis functions**

The functions discussed above can be used for producing weighted estimates, but they do not allow users to fully account for survey design features such as stratification, clustering and unequal selection probabilities.

The `survey` package provides a set of functions that apply survey weights and correctly account for the sample design when estimating standard errors and confidence intervals. Before using the package, you must identify the survey weights and design variables and create a survey design object. This resource includes step-by-step examples showing how to do this.

The `srvyr` package is based on the `survey` package and designed to work better with `dplyr` and the tidyverse approach to R including piping.

# Data and software requirements

To work through the examples on your own machine, you will need:

- basic familiarity with R.

- an up-to-date copy of R.

- the `dplyr`, `haven`, `survey`, `Hmisc` and `ggplot2` packages.

- a UK Data Service account (for downloading the data)

- a copy of the [2017 British Social Attitudes Survey (BSA)](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=8450) - the examples use the Stata (.dta) version of the dataset.

# Identifying the survey design

From the information in the dataset user guide, we can see that the 2017 British Social Attitudes (BSA) survey used a three-stage stratified random sample.

- Postcode sectors were selected first, followed by addresses and then individuals.

- The primary sampling units (postcode sectors) were:

  - first stratified by sub-region, population density, and the proportion of owner-occupied housing.

  - selected with probability proportional to size, so postcode sectors containing more addresses had a greater chance of being selected.

To account for this design in the analysis, we need to identify the variables that contain information on the sampling weights, the strata, and the primary sampling units (clusters).

From the data documentation, we can find the following.

| Survey feature        | Variable   |
|-----------------------|------------|
| Weight                | `WtFactor` |
| Stratum               | `StratID`  |
| Primary sampling unit | `Spoint`   |

# Preparing the data

The first step is to load the R packages we need and to read in the data. Note the code uses the Stata (.dta) version of the dataset. To run the code, you will need to update the file path to match the location where you saved the downloaded data.

```{r}

# Load required packages
library(dplyr)   # Data manipulation
library(haven)   # Import Stata/SPSS files
library(Hmisc)   # Weighted statistics
library(survey)  # Survey design
library(ggplot2) # Visualisation

# Read in the data 

# ------------------------------------------------------------------
# DATA LOCATION: 
# Save the UKDA-8450-stata folder somewhere on your computer.
# Then edit the path below to match its location.
# Example:
# data_dir <- "C:/Data/UKDA-8450-stata"
# ------------------------------------------------------------------

data_dir <- "C:/Data/UKDA-8450-stata"


# Build the path to the dataset

bsa_file <- file.path(
  data_dir,
  "stata",
  "stata13",
  "bsa2017_for_ukda.dta"
)

# Import the dataset using Haven
bsa17 <- read_dta(bsa_file)

# Check that the import was successful
dim(bsa17)


```

The dataset should contain 3988 cases (rows) and 580 variables (columns).

# Example 1: Mean age

This example shows how weighting and survey design can affect survey estimates. We will:

- calculate the unweighted mean age of respondents and compare it with a weighted estimate (using `Hmisc`).

- construct an approximate confidence interval using a general weighting function.

- use the `survey` package to estimate the mean age and its confidence interval while accounting for the survey design.

## Unweighted and weighted mean age

To start, we will first calculate the unweighted mean age.

```{r}

# Unweighted mean age

mean_age <- mean(
  bsa17$RAgeE,
  na.rm = TRUE
)

mean_age

```

To calculate the weighted mean, we first use `wtd.mean()`, from the `Hmisc` package.

```{r}

# Weighted mean age
weighted_mean_age <- wtd.mean(
  bsa17$RAgeE,
  weights = bsa17$WtFactor,
  na.rm = TRUE
)

weighted_mean_age

```

How do they compare?

```{r}
#| echo: false
# Display results

cat("Unweighted mean age:", round(mean_age, 1), "years\n")
cat("Weighted mean age:", round(weighted_mean_age, 1), "years\n")
difference <- weighted_mean_age - mean_age
cat(
  "Difference:",
  round(difference, 1),
  "years\n"
)
```

In this example, the weighted and unweighted estimates differ by around four years. This suggests that older respondents are over-represented in the unweighted sample relative to the target population. Applying the survey weights reduces this imbalance and produces a lower estimate of the population mean age.

## Approximate weighted confidence interval

We can calculate a confidence interval for the weighted mean age. The confidence interval reflects the uncertainty around the estimate. To do this, we first estimate a weighted standard error using the Hmisc package and then use it to construct an approximate 95% confidence interval.

The approximate confidence interval is calculated as: *mean ± 1.96 × standard error.*

Because this approach does not account for stratification or clustering, the confidence interval is an approximation.

```{r}
# APPROXIMATE 95% CONFIDENCE INTERVAL  

# Calculate the weighted standard error (approximate)
weighted_se_age <- sqrt(
  wtd.var(
    bsa17$RAgeE,
    weights = bsa17$WtFactor,
    normwt = TRUE,
    na.rm = TRUE
  ) / sum(!is.na(bsa17$RAgeE))
)

# Calculate the lower and upper confidence limits
lower_ci <- weighted_mean_age - 1.96 * weighted_se_age
upper_ci <- weighted_mean_age + 1.96 * weighted_se_age

# Display the results
cat(
  "Approximate 95% confidence interval:",
  round(lower_ci, 1),
  "to",
  round(upper_ci, 1),
  "years\n"
)
```

The approximate 95% confidence interval suggests that the population mean age lies between 47.7 and 48.9 years. Later, we will compare this interval with one that accounts for the survey design.

## Specifying the survey design in R

The `survey` package provides functions that apply survey weights and account for the survey design when estimating standard errors and confidence intervals.

To use these functions, we first need to specify the survey design by creating a survey design object. A survey design object combines the survey dataset with information about the survey weights and sample design. I

To create a survey design object, we need the variables identified earlier:

- `Spoint` - the primary sampling unit identifier.
- `StratID` - the stratum identifier.
- `WtFactor` - the survey weight.

We create a survey design object using `svydesign().` The resulting object is then used with functions from the `survey` package to produce estimates that account for the survey design.

```{r}
# Create a survey design object
bsa17.s<-svydesign(ids=~Spoint,
                   strata=~StratID,                     
                   weights=~WtFactor,                    
                   data=bsa17)    
```

The survey design object is called `bsa17.s.`

## Using the survey package

We can now use functions from the `survey` package to produce estimates that account for the survey design. Here we use `svymean()` to estimate mean age of respondents.

```{r mean}
#Estimate mean age using the svymean() from the survey package

survey_mean_age <- svymean( 
  ~RAgeE,                   
  design = bsa17.s, # the survey design object
  na.rm = TRUE
)
survey_mean_age
```

The output includes the weighted mean and its standard error. We can use `confint()` to calculate a 95% confidence interval for the estimated mean.

```{r}
# use confint to calculate a confidence interval

survey_ci <- confint(survey_mean_age)
survey_ci

```

We can now compare the survey-based estimate and confidence interval with the weighted estimate and approximate confidence interval calculated earlier.

```{r}
#| echo: false

comparison <- round(
  rbind(
    "Weighted estimate (Hmisc)" = c(
      Mean = weighted_mean_age,
      Lower = lower_ci,
      Upper = upper_ci
    ),
    "Survey estimate" = c(
      Mean = coef(survey_mean_age),
      Lower = survey_ci[1],
      Upper = survey_ci[2]
    )
  ),
  1
)

colnames(comparison) <- c(
  "Mean age",
  "95% CI lower",
  "95% CI upper"
)

knitr::kable(
  comparison,
  caption = "Comparison of weighted mean age estimates and confidence intervals"
)
```

The weighted mean age looks identical using both approaches. However, the confidence interval produced by the survey package is slightly wider than the approximate confidence interval. This is because the survey package accounts for clustering and stratification when estimating uncertainty, whereas the approximate method only uses the survey weights.

## Key points

- Applying survey weights reduced the estimated mean age from 52.2 years to 48.3 years.
- The weighted mean was virtually identical whether calculated using `Hmisc` or the `survey` package.
- The `survey` package produced a wider confidence interval because it accounts for the sample design when estimating uncertainty.

# Example 2: Estimating a proportion

In this example, we will use the `survey` package and our survey design object to estimate the proportion of respondents who report being interested in politics.

The variable is `Politics`. Before calculating proportions, let's examine the distribution of the variable and its response categories.

```{r}
# Frequency distribution of political interest
politics_freq <- table(
  droplevels(
    as_factor(
      bsa17$Politics,
      levels = "both"
    )
  )
)
# Display frequencies in a table
knitr::kable(
  as.data.frame(politics_freq),
  col.names = c("Response", "Frequency")
)
```

Categories `1` ("A great deal") and `2` ("Quite a lot") represent respondents who report relatively high levels of interest in politics. In the next step, we combine these categories to estimate the proportion of adults who are interested in politics (in 2017).

```{r 5_5}

# Estimate the proportion interested in politics

politics_interest <- svyciprop(       # function for proportion 
  ~I(Politics == 1 | Politics == 2),  # for those who answered 1 or 2
  bsa17.s                             # survey design object
)
politics_interest                     # print result
```

The output from `svyciprop()` gives the estimate as a proportion. For easier interpretation, we can convert the estimate and confidence interval to percentages.

```{r}
# Convert estimate and confidence interval to percentages

round(
  data.frame(
    Estimate = coef(politics_interest),
    Lower = confint(politics_interest)[1],
    Upper = confint(politics_interest)[2]
  ) * 100,
  1
)
```

**Try it:**

Use svyciprop and the variable RAgecat5 to estimate the proportion of respondents aged 17-35. Run the code and compare to the solution below.

::: {.callout-note collapse="true"}
## Solution

```{r}

# Explore the age-group variable
table(
  droplevels(
    as_factor(
      bsa17$RAgecat5,
      levels = "both"
    )
  )
)

# Use svyciprop() to estimate the proportion and confidence interval
young_prop <- svyciprop(
  ~I(RAgecat5 == 1),     
  bsa17.s            
)

# Convert the estimate and confidence interval to percentages
round(
  data.frame(
    Estimate = coef(young_prop),
    Lower = confint(young_prop)[1],
    Upper = confint(young_prop)[2]
  ) * 100,
  1
)
```
:::

## Example 2: Key points

- `svyciprop()` can be used to estimate proportions and confidence intervals while accounting for the survey design.

- The group of interest is specified using a logical expression, for example `~I(RAgecat5 == 1)`.

- Estimates are returned as proportions but can be converted to percentages for easier interpretation.

# Example 3: Estimates for population sub-groups (domain estimates)

In Example 2, we estimated the proportion of adults who reported being interested in politics. In this example, we will extend this analysis by estimating the proportion interested in politics separately within each Government Office Region (`GOR_ID`). These are examples of **domain estimates** because they describe subgroups of the population. In the `survey` package, one way of producing domain estimates is using `svyby().`

In this example, we will:

- create a new binary variable `politics_interest` to capture those interested in politics.

- recreate the survey design object so that it contains the new variable.

- use `svyby()` and `svymean` to estimate the proprtoin interested in each Goverment Office Region.

- convert the proportions to percentages.

- plot the percentages to make it easier to compare.

```{r 5.8.2}

# Create indicator: interested in politics = 1, otherwise 0
bsa17$politics_interest <- ifelse(
  bsa17$Politics %in% c(1, 2),
  1,
  0
)

# Recreate survey design object so it includes the new variable
bsa17.s <- svydesign(
  ids = ~Spoint,
  strata = ~StratID,
  weights = ~WtFactor,
  data = bsa17
)

# Estimate percentage interested in politics by region
politics_region <- svyby(
  ~politics_interest,
  ~as_factor(GOR_ID),
  design = bsa17.s,
  FUN = svymean,
  vartype = "ci",
  na.rm = TRUE
)

# Convert estimates and CIs to percentages
politics_region[, 2:4] <- politics_region[, 2:4] * 100

# Round numeric columns only
politics_region[, 2:4] <- round(politics_region[, 2:4], 1)

# print region, estimate and confidence intervals
politics_region [2:4] 
```

Let's now plot the point estimates and their confidence intervals.

```{r}
plot_region <- politics_region

names(plot_region)[1:4] <- c(
  "region",
  "percent_interested",
  "ci_lower",
  "ci_upper"
)

ggplot(
  plot_region,
  aes(x = percent_interested, y = region)
) +
  geom_point() +
  geom_errorbar(
    aes(xmin = ci_lower, xmax = ci_upper),
    orientation = "y",
    width = 0.2
  ) +
  labs(
    x = "Interested in politics (%)",
    y = "Government Office Region"
  )
```

The estimates show political interest varies across regions, ranging from around one-third of adults in the North East to just over one-half in London. The confidence intervals indicate that some of this variation may reflect sampling uncertainty.

## An additional note on subsetting

Another way to estimate statistics for a subgroup is to analyse only respondents belonging to that group. For example, if we were interested only in respondents who are employed, we could create a subset and calculate estimates for that subgroup.

There are a few points to note when subsetting survey data.

- Survey weights are usually designed to make the full sample representative of the target population. If we analyse only a subset of respondents, the weights may not necessarily produce estimates that are representative of the corresponding population subgroup.

- It is important to retain all the survey design information when estimating standard errors and confidence intervals. If subsetting, you should create the survey design object first and then subset using the `subset()` function. This ensures that the survey weights, clustering and stratification information are retained in the analysis.

Although subsetting can be useful when analysing a subgroup, domain estimation using `svyby()` is generally the preferred approach because it is simpler, less prone to error and naturally retains the survey design information.

## Key points

- Domain estimates describe subgroups of the population, such as regions, age groups or men and women.

- `svyby()` can be used to estimate statistics for multiple subgroups in a single command while accounting for the survey design.

- If new variables are added to the dataset after creating a survey design object, the survey design object should be recreated before analysis.

- Be cautious when analysing a subgroup separately. Survey weights may not make the subsample representative of the corresponding population subgroup. Always create the survey design object first and then use subset() to retain all the survey design information.
