---
title: Using survey weights and survey design variables
subtitle: R examples
institute: UK Data Service

date: "June 2026"
date-format: MMMM YYYY
format:
  html:
    css: ukds_long.css
    embed-resources: false
    toc: true
    toc-depth: 2
    number-sections: true

execute:
  echo: true
  warning: false
  message: false
---

- This resource shows how to apply survey weights and account for survey design in R. Using data from the 2017 British Social Attitudes Survey, we will see how weighting changes estimates, how survey design can be specified in R, and why accounting for the sample design matters when assessing uncertainty.

- For an introduction to the concepts covered here, including survey weights, stratification, clustering and design variables, see the companion resource *Survey explainer: Survey weights and design variables*.

- Survey weights help make survey estimates more representative of the target population. Many surveys also use complex sample designs involving stratification and clustering. These design features should be taken into account when estimating standard errors and confidence intervals.

# Survey weighting and design in R

Base R includes only limited support for weighted analysis. However, several R packages provide functions for calculating weighted estimates.

For example, the `Hmisc` package includes functions such as `wtd.mean()`, `wtd.var()` and `wtd.quantile().`The `summarytools` package supported weighted frequncies tables and common descriptive statistics. The functions from these packages can be used for producing weighted estimates, but they do not all you to fully account for survey design features such as stratification, clustering and unequal selection probabilities.

The `Survey` package provides a set of functions that apply survey weights and correctly account for the sample design when estimating standard errors and confidence intervals.

To use the Survey package, you need to:

- Identify the survey weights and design variables.
- Create a survey design object via the command `svydesign` - this tells R how the survey was designed and which weights to use.
- Compute estimates with survey-specific functions.

# Data and software requirements

To work through the examples on your own machine, you will need:

- Basic familiarity with R.

- An up-to-date copy of R.

- The `dplyr`, `haven`, `survey`, `Hmisc` and `ggplot` packages.

- A UK Data Service account (for downloading the data)

- A copy of the [2017 British Social Attitudes Survey (BSA)](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=8450) - the examples use the Stata (.dta) version of the dataset.

# Identifying the survey design

The 2017 British Social Attitudes (BSA) survey used a three-stage stratified random sample. Postcode sectors were selected first, followed by addresses and then individuals.

The primary sampling units (postcode sectors) were:

- stratified by sub-region, population density, and the proportion of owner-occupied housing.

selected with probability proportional to size, so postcode sectors containing more addresses had a greater chance of being selected.

To account for this design in the analysis, we need to identify the variables that contain information on the sampling weights, the strata and the primary sampling units (clusters).

From the documentation that comes with the survey, we can find the following.

| Survey feature        | Variable   |
|-----------------------|------------|
| Weight                | `WtFactor` |
| Stratum               | `StratID`  |
| Primary sampling unit | `Spoint`   |

# Preparing the data 

```{r}

# Load required packages
library(dplyr)   # Data manipulation
library(haven)   # Import Stata/SPSS files
library(Hmisc)   # Weighted statistics
library(survey)  # Survey design
library(ggplot2) # Visualisation

# ------------------------------------------------------------------
# DATA LOCATION
# ------------------------------------------------------------------
# Save the UKDA-8450-stata folder somewhere on your computer.
# Then edit the path below to match its location.
#
# Example:
# data_dir <- "C:/Data/UKDA-8450-stata"

data_dir <- "C:/Data/UKDA-8450-stata"

# Check that the data directory exists
if (!dir.exists(data_dir)) {
  stop(
    "Data directory not found. Please edit 'data_dir' so it points to your UKDA-8450-stata folder."
  )
}

# ------------------------------------------------------------------
# IMPORT DATA
# ------------------------------------------------------------------

bsa_file <- file.path(
  data_dir,
  "stata",
  "stata13",
  "bsa2017_for_ukda.dta"
)

# Check that the dataset exists
if (!file.exists(bsa_file)) {
  stop(
    paste(
      "Dataset not found:",
      bsa_file,
      "\nCheck that 'data_dir' points to the correct folder."
    )
  )
}

# Import Stata dataset
bsa17 <- read_dta(bsa_file)

# Check dimensions
dim(bsa17)

# View the dataset in RStudio
# View(bsa17)
```

# Example 1: Mean age

To look at the effect of weighting and accounting for survey design, we are going to compare estimates for mean age. To start, we will first calculate the unweighted mean age. We will then compare this with a weighted estimate.

## Unweighted mean age

```{r}
# ------------------------------------------------------------------
# UNWEIGHTED MEAN AGE
# ------------------------------------------------------------------

mean_age <- mean(
  bsa17$RAgeE,
  na.rm = TRUE
)

mean_age

```

## Weighted mean age

To calculated the weighted mean we first use `wtd.mean()`, from the `Hmisc` package.

```{r}

# Weighted mean age
weighted_mean_age <- wtd.mean(
  bsa17$RAgeE,
  weights = bsa17$WtFactor,
  normwt = TRUE,
  na.rm = TRUE
)

weighted_mean_age

```

### How do the means compare?

```{r}
#| echo: false
# Display results

cat("Unweighted mean age:", round(mean_age, 1), "years\n")
cat("Weighted mean age:", round(weighted_mean_age, 1), "years\n")
difference <- weighted_mean_age - mean_age
cat(
  "Difference:",
  round(difference, 1),
  "years\n"
)
```

In this example, the weighted and unweighted estimates differ. The difference suggests that older responsents are over-represented in the unweighted sample relative to the population, and the survey weights adjust for this imbalance.

### Approximate weighted confidence interval

We can also add confidence intervals to calculate the precision of our estimate of mean age. At this point, we are not able to account for survey design, the confidence intervals are approximate. Later we will compare with confidence interval we get once accounting for survey design.

```{r}
# ------------------------------------------------------------------
# APPROXIMATE 95% CONFIDENCE INTERVAL
# ------------------------------------------------------------------

weighted_se_age <- sqrt(
  wtd.var(
    bsa17$RAgeE,
    weights = bsa17$WtFactor,
    normwt = TRUE,
    na.rm = TRUE
  ) / sum(!is.na(bsa17$RAgeE))
)

lower_ci <- weighted_mean_age - 1.96 * weighted_se_age
upper_ci <- weighted_mean_age + 1.96 * weighted_se_age

cat(
  "Approximate 95% confidence interval:",
  round(lower_ci, 1),
  "to",
  round(upper_ci, 1),
  "years\n"
)
```

### Specifying the survey design in R

To use the R Survey package we need to:

- use the survey weights and design variables

  - `Spoint` as the primary sampling unit identifier;
  - `StratID` as the stratum identifier;
  - `WtFactor` as the survey weight.

- create a survey design object via the command `svydesign` - this tells R how the survey was designed and which weights to use.

- Compute estimates with survey-specific functions.

### Create the survey design object

```{r}
# Create a survey design object
bsa17.s<-svydesign(ids=~Spoint,
                   strata=~StratID,                     
                   weights=~WtFactor,                    
                   data=bsa17)    
```

### Using the survey package

Now we use the command `svymean()` from the survey package.

```{r mean}
#Estimate mean age using the survey package

survey_mean_age <- svymean(
  ~RAgeE,
  design = bsa17.s,
  na.rm = TRUE
)
survey_mean_age
```

The function also gives the standard error. We can use confint to calculate a confidence interval for the estimated mean.

```{r}
survey_ci <- confint(survey_mean_age)

survey_ci

```

Let's now compare with the estimates we obtained earlier.

```{r}
# Compare weighted mean estimates

round(
  rbind(
    Hmisc = c(
      Mean = weighted_mean_age,
      Lower = lower_ci,
      Upper = upper_ci
    ),
    Survey = c(
      Mean = coef(survey_mean_age),
      Lower = survey_ci[1],
      Upper = survey_ci[2]
    )
  ),
  1
)
```

The weighted mean age is virtually identical using both approaches. However, the confidence interval produced by the survey package is slightly wider than the approximate confidence interval. This is because the survey package accounts for clustering and stratification when estimating uncertainty, whereas the approximate method only uses the survey weights.

> Key points
>
> - Applying survey weights reduced the estimated mean age from 52.2 years to 48.3 years.
> - The weighted mean was virtually identical whether calculated using `Hmisc` or the `survey` package.
> - The `survey` package produced a wider confidence interval because it accounts for the sample design when estimating uncertainty.

# Example 2: Estimating a proportion

The same approach can be used to estimate proportions. In this example, we will estimate the proportion of respondents who report being interested in politics. The variable is `Politics`. It has five categories ranging from 1: 'A great deal' to 5: 'Not at all'. We define respondents as interested in politics of they answered either *A great deal* or *Quite a lot*.

```{r}

# Frequency distribution of political interest
table(
  droplevels(
    as_factor(bsa17$Politics)
  )
)

```

Estimate the percentage of those who were interested in politics in 2017

```{r 5_5}
  # Estimate the proportion interested in politics

politics_interest <- svyciprop(
  ~I(Politics == 1 | Politics == 2),
  bsa17.s
)

round(
  100 * coef(politics_interest),
  1
)
```

Try it yourself: Use svyciprop and the variable RAgecat5, to estimate the proportion of respondents aged 17-35.

::: {.callout-tip collapse="true"}
## Show solution

```{r}
young_prop <- svyciprop(
  ~I(RAgecat5 == 1),
  bsa17.s
)

round(
  100 * coef(young_prop),
  1
)

round(
  100 * confint(young_prop),
  1
)
```
:::

## Example 3: Estimates for population sub-groups (domain estimates)

In Example 2, we estimated the proportion of adults who reported being interested in politics. We can extend this analysis by estimating the proportion interested in politics separately within each Government Office Region.

These are examples of domain estimates because they describe subgroups of the population.

However, when we want to compute estimates for a sub-group of the population, we need to remember that weights are usually designed to make the full sample representative of the population.

We could creation regional datasets to analyse respondents belonging in each subgroup. However, this can lead to incorrect estimates of uncertainity because information from the rest of the sample is discarded when estimating standard errors.

A better approach is to use survey functions that estimate statistics for subgroups while reatining information from the full durvey design. In the survey packages, this can be done using functions such as `svyby().`

## Q5 Answer (1)

Proportion of those interested in politics by region:

```{r 5.8.2}

# Create indicator: interested in politics = 1, otherwise 0
bsa17$politics_interest <- ifelse(
  bsa17$Politics %in% c(1, 2),
  1,
  0
)

# Recreate survey design object so it includes the new variable
bsa17.s <- svydesign(
  ids = ~Spoint,
  strata = ~StratID,
  weights = ~WtFactor,
  data = bsa17
)

# Estimate percentage interested in politics by region
politics_region <- svyby(
  ~politics_interest,
  ~as_factor(GOR_ID),
  design = bsa17.s,
  FUN = svymean,
  vartype = "ci",
  na.rm = TRUE
)

# Convert estimates and CIs to percentages
politics_region[, 2:4] <- politics_region[, 2:4] * 100

# Round numeric columns only
politics_region[, 2:4] <- round(politics_region[, 2:4], 1)

politics_region
```

## Plotting CIs

Let's plot the output to help us compare both the point estimates and the the uncertainity.

```{r}
plot_region <- politics_region

names(plot_region)[1:4] <- c(
  "region",
  "percent_interested",
  "ci_lower",
  "ci_upper"
)

ggplot(
  plot_region,
  aes(x = percent_interested, y = region)
) +
  geom_point() +
  geom_errorbar(
    aes(xmin = ci_lower, xmax = ci_upper),
    orientation = "y",
    width = 0.2
  ) +
  labs(
    x = "Interested in politics (%)",
    y = "Government Office Region"
  )
```

Political interest varies across regions, ranging from around one-third of adults in the North East to just over one-half in London. The confidence intervals indicate that some of this variation may reflect sampling uncertainty, particularly in regions with smaller sample sizes.
