---
title: Using survey weights and survey design variables
subtitle: R examples
institute: UK Data Service

bibliography: inference.bib

date: "June 2026"
date-format: MMMM YYYY
format:
  html:
    css: ukds_long.css
    embed-resources: true
    toc: true
    toc-depth: 2
    number-sections: true

execute:
  echo: true
---

## Data and software requirements

- An up to date copy of R (with the `dplyr`, `haven`, `survey` and `Hmisc` packages installed) or Stata

- An active account with the UK Data Service (for downloading the datasets)

- We will use data from:

  - The [2017 British Social Attitudes Survey (BSA)](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=8450)[@NatCen2024]

  - The [April-June 2022 Quarterly Labour Force Survey](https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=8999#!/access-data)[@ONS2025]

## R

- General

  - R has sime general options for weighting data. For example, base R includes `weighted.mean() and`some R packages offer a more comprehensive set of functions:

    - `Hmisc`: `wtd.mean()`, `wtd.var()`, `wtd.quantile()`
    - `summarytools`: `descr()`, `freq()`

- Survey specific: The R *Survey* package [@Lumley2024] provides a comprehensive set of functions for computing point and variance estimates from survey data.

- Overall logic:

  - Install the `survey` packages and load it into memory.
  - Declare the survey design: i.e. create a `svydesign` object from the data.
  - Compute estimates with survey-specific functions.

### Stata

- Stata provides comprehensive support for computing survey design-informed estimates from survey data

- Stat's SVY

  - Declare the survey design using `svyset`

    - `svyset id=psu_var [pweights=weight_var],strata(strata_var)`

  - Subsequently use `svy:` - prefixed commands for estimation:

    - `svy:mean myvar`, `svy:tab myvar` etc...

## Stata

- Stata offers a general option for usign weights. You need to identify the type of weight from a choice of four kinds of weight:

  - frequency weights (`fweight`),
  - analytical weights (`aweight`),
  - importance weights (`iweight`) and
  - probability weights (`pweight`).

- The options compute variance (and therefore standard errors) in different ways.

- Survey weights should be treated as *probability weights* or `pw`.

- To get a weight estimate, you specify the weight in square brackets w.g.

  `stata_command myvar [pw=weight].`

- However, some common commands such as `summarise` or `tab` do not allow `pw` . While it is possible to use an alternative weighting function instead (`fw` or `aw`) , it is not ideal. You are likely to get correct point estimates but variance/standard errors are likely to be incorrect.

## Identifying the survey design

2017 BSA - learn about the survey design in the documentation that comes together with the data under the `mrdoc/pdf` folder.

The 2017 BSA is a three stage stratified random survey, with postcode sectors, addresses, and individuals as the units selected at each stage.

- Primary sampling units were furthermore stratified according to geographies (sub regions), population density, and proportion of owner-occupiers.

- Sampling fraction (or proportion) was proportional to the size of postcode sectors (i.e. number of addresses)

## Finding the survey design variables

- the primary sampling units (sub regions) are identified bySpoint

  - the strata byStratID
  - the weights variable is WtFactor.

- Addresses are not provided but could be approximated with a household identifier.

## Preparing the data - R

```{r}

# Load required packages
library(dplyr)   # Data manipulation
library(haven)   # Import Stata/SPSS files
library(Hmisc)   # Weighted statistics
library(survey)  # Survey design
library(ggplot2) # Visualisation

# ------------------------------------------------------------------
# DATA LOCATION
# ------------------------------------------------------------------
# Save the UKDA-8450-stata folder somewhere on your computer.
# Then edit the path below to match its location.
#
# Example:
# data_dir <- "C:/Data/UKDA-8450-stata"

data_dir <- "C:/Data/UKDA-8450-stata"


# ------------------------------------------------------------------
# IMPORT DATA
# ------------------------------------------------------------------

bsa_file <- file.path(
  data_dir,
  "stata",
  "stata13",
  "bsa2017_for_ukda.dta"
)


# Check dimensions
dim(bsa17)

# View the dataset in RStudio
# View(bsa17)
```

Mean

## Mean age

```{r}
mean_age <- mean(bsa17$RAgeE, na.rm = TRUE)

cat("Mean age:", round(mean_age, 1), "years\n")
```

## Weighted mean

```{r}
# ------------------------------------------------------------------
# WEIGHTED MEAN AGE
# ------------------------------------------------------------------

weighted_mean_age <- wtd.mean(
  x = bsa17$RAgeE,
  weights = bsa17$WtFactor,
  normwt = TRUE,
  na.rm = TRUE
)

cat(
  "Weighted mean age:",
  round(weighted_mean_age, 1),
  "years\n"
)
```

```{r}
# ------------------------------------------------------------------
# APPROXIMATE 95% CONFIDENCE INTERVAL
# ------------------------------------------------------------------

weighted_se_age <- sqrt(
  wtd.var(
    x = bsa17$RAgeE,
    weights = bsa17$WtFactor,
    normwt = TRUE,
    na.rm = TRUE
  ) / sum(!is.na(bsa17$RAgeE))
)

lower_ci <- weighted_mean_age - 1.96 * weighted_se_age
upper_ci <- weighted_mean_age + 1.96 * weighted_se_age

cat(
  "95% confidence interval:",
  round(lower_ci, 1),
  "to",
  round(upper_ci, 1),
  "years\n"
)
```

## Specifying the survey design in R

- We can specify the survey design using:
  - `Spoint` as Primary Sampling Unit ids,
  - `StratID` as strata ids,
  - `WtFactor` as weights.
- We create a `svydesign` object. A survey design informed copy of the data. which will be used for subsequent estimation.

```{r}

bsa17.s <- svydesign(
  ids = ~Spoint,
  strata = ~StratID,
  weights = ~WtFactor,
  data = bsa17
)

summary(bsa17.s)          
```

## Mean age and its 95% CI

Estimate mean age of respondents using `svymean()`

```{r mean}
svymean(
  ~RAgeE,
  design=bsa17.s
  na.rm=TRUE)
```

By default `svymean()` computes the standard error of the mean. We can use `confint()` to also get a confidence interval.

```{r 5_3}
confint(                                  ### Computing the confidence interval...
  svymean(~RAgeE,bsa17.s)                 ### And the mean
  )                   
```

- For a slightly nicer output:

```{r nicer1}
#| code-fold: true
round(                                    ### Rounding the results to one decimal point
  c(
    svymean(~RAgeE,bsa17.s),              ### Computing the mean...
    confint(svymean(~RAgeE,bsa17.s))      ### And its 95% CI
    ),
  1) 
```

## Example general weighting

```{r}
a.m<-wtd.mean(                               ### Weighted mean function from the
  bsa17$RAgeE,                               ### Hmisc package
  bsa17$WtFactor,
  normwt = T)                                ### Option specific to survey weights  

### Computation of the standard error by hand...
a.se<-sqrt(
          wtd.var(bsa17$RAgeE,            ### ... using the weighted variance function from Hmisc
                  bsa17$WtFactor,
                  normwt = T)
          )/
       sqrt(
         nrow(bsa17)                         ### ... shortcut to sample size
         )

c(a.m,                                       ### Concatenating the mean..
  a.m-1.96*a.se,                             ### ... the lower bound of the CI
  a.m+1.96*a.se)                             ### ... and the higher bound
```

## Unweighted estimate

```{r}
ua.m<-mean(bsa17$RAgeE)                     ### mean() function from R Base
  
ua.se<-sd(bsa17$RAgeE)/                     ### ... standard errors
      sqrt(nrow(bsa17))    ##

c(ua.m,                                     ### and putting it all together
  ua.m-1.96*ua.se,
  ua.m+1.96*ua.se
  )

```

## Summary

Not using weights results in overestimating the mean age in the population (of those aged 18+) by about 4 years. This might be due to the fact that older respondents are more likely to take part to surveys.

Using a general weighting command did not alter the value of the estimated population mean when compared with SD informed estimates.

... but would lead us to overestimating the precision/underestimate the uncertainty of our estimate -- by about plus or minus 3 months.

## Computing a proportion and its 95% confidence interval

- We can similarly estimate the distribution of a categorical variable in the population by estimating proportions (or percentages)
- Let's look at the proportion of people who declare that they are interested in politics.
- This is the `Politics` variable in the BSA.
- It has five categories ranging from 1: 'A great deal' to 5: 'Not at all'.
- We could recode together 1 and 2, i.e. `A great deal` and `quite a lot` into `Significantly`, but here we will directly select the relevant values on the fly as this is allowed by R.

## Let's explore the variable

- Phrasing of the question:

```{r 5_4_1}
attr(bsa17$Politics,"label")     
```

- Sample distribution

```{r 5_4_3}
table(
  droplevels(                                    ### We are using droplevels() 
             as_factor(bsa17$Politics, "both")   ### in order to hide categories
             )                                   ### ... without any observations
  ) 
```

## 

- Let's infer the percentage of those who were interested in politics in 2017

```{r 5_5}
  round(100*
    prop.table(
      svytable(~(Politics==1 | Politics==2),bsa17.s)
      ),1)
```

```{r 5_5.1}
#| code-fold: true
#| code-summary: "Alternative:"

bsa17$PolC<-as.factor(ifelse(bsa17$Politics==1 | bsa17$Politics==2, "Interested", "Not so much") )
bsa17.s<-svydesign(ids=~Spoint, 
                   strata=~StratID, 
                   weights=~WtFactor,
                   data=bsa17)        

  round(100*
    prop.table(
      svytable(~(PolC),bsa17.s)
      ),1)


```

- Estimate the confidence intervals for the proportions via `svyciprop()` and `I()`:

  - The former computes the proportion and its confidence interval (by default 95%)...

  - ... the latter allows us to define the category of interest.

```{r 5_6_2}
round(100*
        c("% interested"= p[[1]],                                          ### Extracts the point estimate
          "95% CI - LB"=attr(p,"ci")[[1]],"95% CI - UB"=attr(p,"ci")[[2]]  ### Extracts the CI
          ),1 
      )
```

What is the proportion of respondents aged 17-34 in the sample, as well as its 95% confidence interval? use `RAgecat5.`The proportion of 17-34 year old in the sample is:

```{r}
a<-svyciprop(~I(RAgecat5 == 1),
              bsa17.s)
a
round(
  100*a[1],
     1)
```

and its 95% confidence interval:

```{r}
round(
    100*
    confint(a),  ### Another way of extracting CI from svyciprop objects
    1)
```

## Domain estimates

Computing estimates for subpopulations adds a layer of complexity. Weights are usually designed to make the full sample representative of the population. If we compute estimates for a sub-population only this would mean we are only using a fraction of these weights, which in turn may alter the accuracy of our estimates. It is instead recommended to use commands that take into account the entire distribution of the weights.

- The R command that does this is `svyby()`

## Computing domain estimates in R

- Say we would like to compute the mean age of BSA respondents by Government Office Regions

- We need to specify:

  - The outcome variable whose estimate we want to compute: i.e. `RAgeE`
  - The grouping variable(s) `GOR_ID`
  - The estimation function we are going to use here: `svymean()`
  - And the type of type of variance estimation we would like to see displayed i.e. standard errors or confidence interval

## Output

```{r 5.7}
d<-      svyby(~RAgeE,               ### Outcome variable
             by=~as_factor(GOR_ID),  ### Subpopulations
             svymean,                ### Estimation function
             design=bsa17.s,
             vartype = "ci")         ### CI or SE 
round(d[-1],1)
```

We used `[-1]` above to remove the column with region names from the results, so that we could round the estimates without getting an error.

## Interpretation

- The population in London is among the youngest in the country whereas those in the South West are among the oldest
- This is likely to be statistically significant as their 95% CI do not overlap.
- The same cannot be said of differences between London and the South East, as the CIs partially overlap.

## Q5 Answer (1)

Proportion of those interested in politics by region:

```{r 5.8.2}

c<-svyby(~I(Politics==1 | Politics==2),
            by=~as_factor(GOR_ID),
            svyciprop,
            design=bsa17.s,
            vartype = "ci")

round(100*c[-1],1)
```

## Plotting CIs

Let's now plot this output:

```{r 5.8.3}
#| output: asis
#| fig-alt: "A plot of the proportion of those interested in politics a great deal or quite a lot by regions"

ggplot(c, aes(x = `I(Politics == 1 | Politics == 2)`, y = rownames(c))) +
  geom_point(size = 3, color = "#ff9800") +
  geom_errorbar(aes(xmin = ci_l, xmax = ci_u), width = 0.1, color = "#ff9800") +
  labs(
    title = "% interested in politics by region",
    y = "Government Office Region",
    x = "% Interested in politics 'a great deal'or 'quite a lot'"
  ) +
  theme_minimal(base_size = 14)

```

## Question 5: Answer

- The 95% confidence interval for the proportion of people significantly interested in politics in the North East is `r round(100*as.numeric(c[1,3:4]),1)`.
- By contrast, it is `r round(100*as.numeric(c[7,3:4]),1)` in London.
- The region with the lowest precision of estimates (i.e. the widest confidence interval) is Wales, with more than 20 percentage point difference between the upper and lower bounds of the confidence interval.

## Question 6

- Using interest in politics as before, and three category age `RAgecat5`:
- Produce a table showing the proportion of respondents significantly interested in politics by age group and gender
- Assess whether the age difference in interest for politics is similar for each gender.
- Is it fair to say that men aged under 35 are more likely to declare being interested in politics than women aged 55 and above?

## Question 6 - Answer

```{r 6.1}
round(
      100*
        svyby(~I(Politics==1 | Politics==2),
              by=~as_factor(RAgecat5)+as_factor(Rsex),
              svyciprop,
              design=bsa17.s,
              vartype = "ci")[c(-8,-4),c(-2,-1)],1)
```

- Males and females aged 55+ tend to be more involved in politics than those who are younger.

- CIs for the proportion of men under 35 and women above 55 interested in politics overlap; it is unlikely that they differ significantly in the population.
